Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-01-12 docs/Midyear/Mid_year_report.tex
Submitted Version (maybe)

In this study, we focused on one particular direction: generalize the Erlang distribted stage duration currently achieved in the model to have more diverse shape through elminate the restriction from the shape parameter (K). The value of K in the SInR model also subjected to a set of discrete values.

extending the stage duration distribution inherent in the SIR model to approximate a gamma distribution. Gamma distributions offer a more versatile distribution shape compared to Erlang distributions, as they allow for non-integer shape parameters. The challenge in extending the Linear Chain Trick (LCT) to accommodate Gamma distributions is that their shape parameters cannot be directly decomposed into exponential phases. To overcome this limitation, as well as those previously mentioned, we propose a novel approach. Instead of using a variable number of substages $(n)$ with a constant rate $(\gamma_i = n\gamma)$ for all exponentially distributed substages, we fix $n$ to a specific integer value and use a series of rates that follow a geometric pattern. This series is represented by the formula $\gamma_i = ar^{i-1}$, governed by parameters $a$ and $r$. Consequently, the parameter of interest shifts from discrete values $(n$ and $ng)$ to continuous values $(a$ and $r)$, thereby enhancing the model's adaptability.

To illustrate this point, consider that the Gamma distribution can be parameterized in terms of its mean $(M)$ and quadratic coefficient of variation (also refer as $\kappa$), which determine the overall shape and position of the distribution. As mentioned above, in subdividing a single stage into n substages, each exponentially distributed substage must have a rate of $n\gamma$ to maintain the overall rate of that stage. While the mean of the Erlang distribution in the ODE system can be chosen freely, $\kappa$ is constrained to discrete values, shown as follows:
\begin{align*}
    Mean: \quad &\sum_{i=1}^{n_f} \frac{1}{n\gamma} = \frac{1}{\gamma}\\
    Variance: \quad &\sum_{i=1}^{n_f} \frac{1}{n^2\gamma^2} = \frac{1}{n\gamma^2} \\
    \kappa: \quad &\frac{\sigma^2}{M^2} = \frac{1}{n} \quad n \in \mathbb{N}
\end{align*}
Our approach, using a geometric series of rates, addresses this issue by allowing both the mean and $\kappa$ to take any value. A detailed explanation and the derivation process are provided in section $4.2$. Another advantage of this method is that it enables a fixed model structure where only the parameters change, eliminating the need to determine optimal substage numbers through the construction of multiple models, a process currently required in the ODE model that employs the LCT.

In this study, we primarily focused on the application of the LCT in the epidemic SIR model, which we refer to as the SI$^n$R model, where $n$ represents the number of substages divided. The objective of our study is to develop and evaluate the method of using a geometrically distributed substage rate as a replacement for the constant rate, aiming to enhance flexibility and efficiency. We have named the model that utilizes this method the SIgR model. We hypothesize that this approach will not only be computationally efficient but also provide a better fit to real-world data. In the following sections, we first aim to align the infectious stage duration distribution shapes generated by the SI$^n$R and SIgR models. Once we have confirmed that our proposed model can effectively substitute the SInR model, we proceed to allow the SIgR model to generate stage duration distributions with characteristics unattainable by the SI$^n$R model. We then invert the process: given a desired mean and $\kappa$ (which can be any real number within a specific range), we develop a computational system to determine the parameter values $(a$ and $r)$ that enable the model to produce a stage duration distribution with these properties. Subsequently, we will conduct empirical validation of our model through data fitting. This evaluation will cover aspects such as the model’s consistency with dataset and its predictive accuracy in forecasting future trends. We anticipate that this approach will enhance the model’s adaptability and practicality, thus paving the way for the development of user-friendly software tools that empower modelers to efficiently select and employ the most suitable models for describing and predicting future epidemic scenarios.

@@

The idea of dividing the infectious stage into $n$ sub-stages exploits the fact that the sum of a sequence of independent, exponentially distributed random variables follows an Erlang distribution (i.e. gamma distributed with integer shape parameter) [Therrien and Tummala 2011]. Thus, we begin by confirming the uniformity of the results generated by simulating the SI$^n$R model with the Erlang distribution, which has the following form:
\begin{equation*}
    f(x; n,n\gamma) = \frac{(n\gamma)^n}{(n-1)!} x^{n-1} e^{-n\gamma x} \quad x>0, n \in \mathbb{N}
\end{equation*}

@@

We aim to fix the model structure (number of substages) and alter the parameter $a$ and $r$ to determine if this approach is more effective than changing the number of substages. Our starting point involves keeping the mean stage duration $(M)$ constant while varying $r$ to explore the possibility of a match between the SI$^n$R and SIgR models. This match will be assessed both visually and by the shape parameter, which is the coefficient of variation squared $(\kappa = \sigma^2)$.

After determining the potential match, we have formulated the formulas for $M$ and $\kappa$ in terms of $a$ and $r$. Utilizing the geometric distribution of rates in the ODE model allows for independent changes in $M$ and $\kappa$. This enables us to achieve the desired value of $\kappa$ without altering the mean, and vice versa, a flexibility not present in the traditional SI$^n$R model. Our goal is to reverse the method: starting with given values of $M$ and $\kappa$, we aim to find specific values of $a$ and $r$ that will let the model generate the stage duration distribution that satisfies these properties. Due to the complexity introduced by the high powers in the formula, solving it explicitly is challenging. To address this, we have developed a system of functions in the R programming language, employing the uniroot() function to obtain a numerical solution for r. We then reintegrate this parameter value into the SIgR model and compare it with the SInR model, using the same $M$ and $\kappa$ values. Subsequently, the mean and the shape parameter of both simulation results are calculated numerically to confirm the validity of the match.

After confirming that the newly proposed model can effectively replace the SInR model by encompassing the same infectious stage duration distributions, we will proceed to empirically test the enhanced performance of the model through data fitting. This process will be divided into two phases and will be conducted in R using the \textbf{bbmle} package. As the starting point, we will employ a deterministic model and assume only observation noise. Our first step involves coding to generate a 'fake' dataset, treating the disease incidence time series from the SIgR model as the actual case, and introducing deviations (noises) into them using a mean reporting ratio and a negative binomial dispersion parameter. The model parameters will then be estimated using the Maximum Likelihood Estimation (MLE) method, applied to the fake dataset. The objective here is to accurately recover the original parameter values that were used to generate the disease incidence data. In the next phase of the study, we will fit the model using an real-world dataset. We plan to compare the fitting results obtained from both the SI$^n$R and SIgR models. This will allow us to quantitatively assess the efficacy of our proposed method in improving the model's accuracy in reflecting actual epidemiological scenarios.

@@

The substage number of the SIgR model $(n_f)$ has been fixed at 12, and the mean stage duration for the entire infectious period is set to 10 days. We examine the effect of using 2, 4, and 8 infectious substages in the SI$^n$R model $(n = 2, 4, 8)$ on the values of $r$ required for the SIgR model to align formulaically with the shape parameter $(\kappa)$. In Figure 2, the plots on the left illustrate the changes in stage duration distribution resulting from varying $r$ within a range that includes the formulaic match for $\kappa$. The black curve represent duration distribution from the SI$^n$R model with the associated $n$. Plots on the right depict the variations in $\kappa$, calculated using the corresponding r values, where the red dashed line is the target value of $\kappa$. It is evident from the plots that, within the specified range of r, there is a concurrence in both the mean and shape parameter values between the SInR and SIgR model. These findings suggest that the results generated by the traditional mode are replicable using the model we propose.

@@

\subsection{Identifying Parameters $a$ and $r$ to Align SI$^n$R and SIgR with Given Mean and $\kappa$}
Utilizing formulas (4) and (6), we have developed a computational system to reverse-engineer the process. The parameters $a$ and $r$, generated by this system, were reintegrated into the SIgR model. This enable us to compare the stage duration distribution with that of the SI$^n$R model, which possesses identical mean and shape parameters, as depicted in Figure 3. Although the formulaic $M$ and $\kappa$ are computed to be the same, we also conducted a numerical calculation of both properties, with results presented in Table 1. This analysis confirms that a match is achieved at both the formulaic and numerical levels. The visual discrepancies observed between the two distributions in Figure xx can be considered negligible, since both distributions serve as theoretical representations of real-world scenarios and do not require exact correspondence.
\begin{figure}[h]
    \centering
    \includegraphics[width= \textwidth]{4.3.1.png}
    \caption{Theoretical Alignment of Mean and Shape Parameters in SI$^n$R and SIgR Models with Calculated Parameter Values $a$ and $r$}
\end{figure}

@@

\section{Discussion}
SIR compartmental models are widely utilized in modeling infectious diseases, but they encounter a significant limitation: the implicit assumption that the time an individual spends in the disease stage follows an exponential distribution. To address this issue, a common method involves subdividing the infectious stage into $n$ number of substages, each substage is exponentially distributed with a constant rate. This approach is known as the Linear Chain Trick (LCT). LCT strategy facilitates an Erlang-shaped stage duration distribution with increased reality. However, this model also faces limitations, including cumbersome parameter optimization processes and the constraint of discrete parameter values, which can affect its performance.

>>>>

In this study, we focused on one particular direction: generalizing the Erlang distribution of stage durations currently present in the model. The goal was to allow for a more diverse range of shapes, achieved by eliminating the constraints imposed by the shape parameter ($\kappa$). The value of $\kappa$ in the SInR model was limited to a set of discrete values, as detailed below. To address this limitation, as well as those previously mentioned, we propose a novel approach. Instead of using a variable number of substages $(n)$ with a constant rate $(\gamma_i = n\gamma)$ for all exponentially distributed substages, we fix $n$ to a specific integer value and use a series of rates that follow a geometric pattern. This series is represented by the formula $\gamma_i = ar^{i-1}$, governed by parameters $a$ and $r$. Consequently, the parameter of interest shifts from discrete values $(n$ and $ng)$ to continuous values $(a$ and $r)$, thereby enhancing the model's adaptability.

To illustrate this point, consider that the Gamma distribution can be parameterized in terms of its mean $(M)$ and shape parameter (quadratic coefficient of variation, $\kappa$), which determine the overall shape and position of the distribution. As mentioned above, in subdividing a single stage into n substages, each exponentially distributed substage must have a rate of $n\gamma$ to maintain the overall rate of that stage. While the mean of the Erlang distribution in the ODE system can be chosen freely, $\kappa$ is constrained to discrete values, shown as follows:
\begin{align*}
    Mean: \quad &\sum_{i=1}^{n_f} \frac{1}{n\gamma} = \frac{1}{\gamma}\\
    Variance: \quad &\sum_{i=1}^{n_f} \frac{1}{n^2\gamma^2} = \frac{1}{n\gamma^2} \\
    \kappa: \quad &\frac{\sigma^2}{M^2} = \frac{1}{n} \quad n \in \mathbb{N}
\end{align*}
Our approach, using a geometric series of rates, addresses this issue by allowing $\kappa$ to take any value within the range $[\frac{1}{n}, 1)$. A detailed explanation and the derivation process are provided in section $4.2$. Another advantage of this method is that it enables a fixed model structure where only the parameters change, eliminating the need to determine optimal substage numbers through the construction of multiple models, a process currently required in the ODE model that employs the LCT.

Throughout this research, we primarily focused on the application of the LCT in the epidemic SIR model, which we refer to as the SI$^n$R model, where $n$ represents the number of substages divided. The objective of our study is to develop and evaluate the method of using a geometrically distributed substage rate as a replacement for the constant rate, aiming to enhance flexibility and efficiency. We have named the model that utilizes this method the SIgR model (temporarily). We anticipate that this approach will not only be computationally efficient but also provide a better fit to real-world data. In the following sections, we first aim to align the infectious stage duration distribution shapes generated by the SI$^n$R and SIgR models. Once we have confirmed that our proposed model can effectively substitute the SInR model, we proceed to allow the SIgR model to include stage duration distributions with characteristics unattainable by the SI$^n$R model. We then invert the process: given a desired mean and $\kappa$, we develop the computational system to determine the parameter values $(a$ and $r)$ that enable the model to produce a stage duration distribution with these properties. Subsequently, we will conduct empirical validation of our model through data fitting. This evaluation will cover aspects such as the model’s consistency with dataset and its predictive accuracy in forecasting future trends. We predict that this approach will enhance the model’s adaptability and practicality, thus paving the way for the development of user-friendly software tools that empower modelers to efficiently select and employ the most suitable models for describing and predicting future epidemic scenarios.

@@

The idea of dividing the infectious stage into $n$ sub-stages exploits the fact that the sum of a sequence of independent, exponentially distributed random variables follows an Erlang distribution (i.e. gamma distributed with integer shape parameter) \cite{therrien2018probability}. Thus, we begin by confirming the uniformity of the results generated by simulating the SI$^n$R model with the Erlang distribution, which has the following form:
\begin{equation*}
    f(x; n,n\gamma) = \frac{(n\gamma)^n}{(n-1)!} x^{n-1} e^{-n\gamma x} \quad x>0, n \in \mathbb{N}
\end{equation*}

@@

We aim to determine if altering the parameter $a$ and $r$ is more effective than changing the number of substages. Our starting point involves keeping the mean stage duration $(M)$ constant while varying $r$ to explore the possibility of a match between the SI$^n$R and SIgR models. This match will be assessed both by the mean $(M)$ and by the shape parameter, which is the coefficient of variation squared $(\kappa = \frac{\sigma^2}{M^2})$.

After determining the potential match, we have formulated the formulas for $M$ and $\kappa$ in terms of $a$ and $r$. Utilizing the geometric distribution of rates in the ODE model allows for independent changes in $M$ and $\kappa$. This enables us to achieve the desired value of $\kappa$ without altering the mean, and vice versa. Our goal is to reverse the method: starting with given values of $M$ and $\kappa$, we aim to find specific values of $a$ and $r$ that will let the model generate the stage duration distribution that satisfies these properties. Due to the complexity introduced by the high powers in the formula, solving it explicitly is challenging. To address this, we have developed a system of functions in the R programming language, employing the uniroot() function to obtain a numerical solution for r. We then reintegrate this parameter value into the SIgR model and compare it with the SInR model, using the same $M$ and $\kappa$ values. Subsequently, the mean and the shape parameter of both simulation results are calculated numerically to confirm the validity of the match.

After confirming that the newly proposed model can effectively replace the SI$^n$R model by encompassing the same infectious stage duration distributions, we will proceed to empirically test the enhanced performance of the model through data fitting. This process will be divided into two phases and will be conducted in R using the \textbf{bbmle} package. As the starting point, we will employ a deterministic model and assume only observation noise. Our first step involves coding to generate a 'fake' dataset, treating the disease incidence time series from the SIgR model as the actual case, and introducing deviations (noises) into them using a mean reporting ratio and a negative binomial dispersion parameter. The model parameters will then be estimated using the Maximum Likelihood Estimation (MLE) method, applied to the fake dataset. The objective here is to accurately recover the original parameter values that were used to generate the disease incidence data. In the next phase of the study, we will fit the model using an real-world dataset. We plan to compare the fitting results obtained from both the SI$^n$R and SIgR models. This will allow us to quantitatively assess the efficacy of our proposed method in improving the model's accuracy in reflecting actual epidemiological scenarios.

@@

The substage number of the SIgR model $(n_f)$ has been fixed at 12, and the mean stage duration for the entire infectious period is set to 10 days. We examine the effect of using 2, 4, and 8 infectious substages in the SI$^n$R model $(n = 2, 4, 8)$ on the values of $r$ required for the SIgR model to align formulaically with the shape parameter $(\kappa)$. In Figure 2, the plots on the left illustrate the changes in stage duration distribution in SIgR model resulting from varying $r$ within a range that includes the formulaic match for $\kappa$. The black curve represent duration distribution from the SI$^n$R model with the associated $n$. Plots on the right depict the variations in $\kappa$, calculated using the corresponding r values, where the red dashed line is the target value of $\kappa$. It is evident from the plots that, within the specified range of r, there is a concurrence in both the mean and shape parameter values between the SInR and SIgR model. These findings suggest that the results generated by the traditional mode are replicable using the model we propose.

@@

\subsection{Identifying Parameters $a$ and $r$ to Align SI$^n$R and SIgR with Given Mean and $\kappa$}
Utilizing formulas (4) and (6), we have developed a computational system to reverse-engineer the process. The parameters $a$ and $r$, generated by this system, were reintegrated into the SIgR model. This enable us to compare the stage duration distribution with that of the SI$^n$R model, which possesses identical mean and shape parameters, as depicted in Figure 3. Although the formulaic $M$ and $\kappa$ are computed to be the same, we also conducted a numerical calculation of both properties, with results presented in Table 1. This analysis confirms that a match is achieved at both the formulaic and numerical levels. The visual discrepancies observed between the two distributions in Figure 3 can be considered negligible, since both distributions serve as theoretical representations of real-world scenarios and do not require exact correspondence.
\begin{figure}[h]
    \centering
    \includegraphics[width= \textwidth]{4.3.1.png}
    \caption{Theoretical Alignment of Mean and Shape Parameters in SI$^n$R and SIgR Models with Calculated Parameter Values $a$ and $r$}
\end{figure}

@@

\section{Discussion}
SIR compartmental models are widely utilized in modeling infectious diseases, but they encounter a significant limitation: the implicit assumption that the time an individual spends in the disease stage follows an exponential distribution. To address this issue, a common method involves subdividing the infectious stage into $n$ number of substages, each substage is exponentially distributed with a constant rate. This approach is also known as the Linear Chain Trick (LCT). LCT strategy facilitates an Erlang-shaped stage duration distribution with increased reality. However, this model also faces limitations, including cumbersome parameter optimization processes and the constraint of discrete parameter values, which can affect its performance.
